Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2025-01-23 paper/main.tex
Updates

% Due to these challenges, whether constructing the model manually or using an
% algorithm, we typically have to rely on some expert knowledge to get a model
% that we can use for further analysis. Hence, it is crucial to have tools to
% guide this construction or modification process.

As a result of these issues, expert knowledge integration is vital to get usable 
results from these algorithms.
% whether constructing the model manually or using
% an algorithm, the user has to rely on their expertise to make modifications to
% the model. 
One way to do this to specify expert knowledge in terms of
fixing edges, blacklisting edges, or specifying tiered knowledge.
The other ways is to do post hoc modification to the learned model using
implied CI based model testing \citep{Ankan2021, Ankan2023}. In this testing
procedure, users test whether the implied CIs of the DAG hold in the dataset or
not using statistical CI tests. Based on the results of these statistical
tests, we can accordingly make modifications to the current model to fix the
violations. However, this CI based testing also has a few drawbacks: 1) CIs are
implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

@@

\section{Background and Related Work}
\label{sec:background}
We consider random variable $ X $ and $ Y $ with a possible multi-dimensional
variable $ \bm{Z} = \{ Z_1, \cdots, Z_k \} $. We consider these variables in
the mixed data setting where these variables can be any combination of
continuous, categorical, or ordinal unless specified. We also assume all
variables to be observed. We use $ \rvert \bm{Z} \rvert $ to represent the
cardinality of the variable $ \bm{Z} $. We write $ \bm{x} = (x_1, \cdots, x_n)
$ for a sample from $ X $ of size $ n $. We write the expectation of a variable
$ X $ as $ \mathbb{E}[X] $, conditional expectations as $ \mathbb{E}[X |
\bm{Z}] $.

>>>>

These challenges highlight the need for methods that effectively combine
automated tools with expert-driven model refinement.  As a result of these
issues, expert knowledge integration is vital to get usable results from these
algorithms. One way to do this to specify expert knowledge in terms of
fixing edges, blacklisting edges, or specifying tiered knowledge.
The other ways is to do post hoc modification to the learned model using
implied CI based model testing \citep{Ankan2021, Ankan2023}. In this testing
procedure, users test whether the implied CIs of the DAG hold in the dataset or
not using statistical CI tests. Based on the results of these statistical
tests, we can accordingly make modifications to the current model to fix the
violations. However, this CI based testing also has a few drawbacks: 1) CIs are
implied only by missing edges in the model, so we can not test for already
existing edges. 2) In high sample size scenarios, p-values can suggest a
significant association between variables even if they have a very weak
relationship. 3) Number of implied CIs can be large making it difficult to
manually go through each of them; for example, in the moderately sized alarm
network with $ 37 $ node, there are $287$ implied CIs.

@@

\section{Background and Related Work}
\label{sec:background}
We consider random variables $ X $ and $ Y $ with a possible multi-dimensional
variable $ \bm{Z} = \{ Z_1, \cdots, Z_k \} $. We consider these variables in
the mixed data setting where these variables can be any combination of
continuous, categorical, or ordinal unless specified. We also assume all
variables to be observed. We use $ \rvert \bm{Z} \rvert $ to represent the
cardinality of the variable $ \bm{Z} $. We write $ \bm{x} = (x_1, \cdots, x_n)
$ for a sample from $ X $ of size $ n $. We write the expectation of a variable
$ X $ as $ \mathbb{E}[X] $, conditional expectations as $ \mathbb{E}[X |
\bm{Z}] $.
